---
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Reading data

Our first exercise entails reading data from files and summarizing and combining the records. This is very common and important step; but somewhat tricky to learn.  We will use the data in a file named "Operating-Health-Facilities-with-location-2014.csv" under "data" read the data in the file into a data.frame and answer all the questions by using R code only. 

Replace the ___________ with the appropriate code 

Feel free to use materials of Introduction to R in: http://rspatial.org/intr/.

1. Read the file into a data.frame.

```{r d1}
d <- read.csv("./data/Operating-Health-Facilities-with-location-2014.csv", stringsAsFactors=FALSE)

```

2. how many rows and columns are there?

```{r d2}
dim(d)
nrow(d)
ncol(d)
```


3. what are the variable names?

```{r d3}
colnames(d)
head(d)
tail(d)
```


4. What is the number of facilities for each zone?
See [data exploration](https://rspatial.org/intr/7-explore.html#data-exploration)

```{r d4}
table(d$ZONE) 
summary(d)
plot(sort(d$LATITUDE))

```

5. Which region has most facilities (and how many), and which has least (and how many)?

```{r d5}
x <- table(d$REGION)
names(which.min(x))
min(x)
names(which.max(x))
max(x)

```

6. What is the average elevation of all clinics?

```{r 6}
x <- d$ALTITUDE[d$FACILITY.TYPE=="Clinic"]
mean(x, na.rm=TRUE)


```


7a. Create a data.frame with four columns: ZONE, REGION, the number of facilities, and their average elevation (by REGION)

> we want something like this

| ZONE      | REGION        | elevation | n   |
|-----------|---------------|-----------|-----|
| 1 Central | Dodoma        | 1059.4    | 369 |
| 2 Central | Singida       | 1355.4    | 210 |
| 3 Eastern | Dar es Salaam | 54.4      | 599 |

We can use the aggregate function. If you struggle, divide and conquer. Do what you can do first. Split the task into parts, and then combine them again later.

```{r 7a}
#Start with mean
a <- aggregate(d[, "ALTITUDE", drop=FALSE], d[, c("ZONE","REGION"), drop=FALSE], mean, na.rm=TRUE)
head(a)
#The number of facilities
b <- aggregate(d[, "FACILITY_ID_NUMBER", drop=FALSE], d[, c("ZONE","REGION"), drop=FALSE], length)
head(b)
a <- merge(a, b, by = c("ZONE","REGION") )

```

Can also be done using a loop as follows:

```{r 7ab}
r <- sort(unique(d$REGION))
result <- matrix(nrow=length(r), ncol=4)
colnames(result) <- c("ZONE", "REGION", "ALTITUDE", "Facilities")
for (i in 1:length(r)) {
  m <- d[d$REGION==r[i], ]
  result[i,1] = unique(m$ZONE)
  result[i,2] = r[i]
  result[i,3] = mean(m$ALTITUDE, na.rm = TRUE)
  result[i,4] = length(m$ALTITUDE)
}
result <- as.data.frame(result)
head(result)

```


7b. provide appropriate column names 

```{r 7b}
colnames(a)[3:4] <- c('ELEVATION', 'FACILITIES No.')
head(a)
```


7c. sort by ZONE, REGION. See Introduction to R. Chapter 11

```{r 7c}
a <- a[order(a$ZONE, a$REGION), ]
head(a)
```


7d. elevation should be rounded to the nearest 10 cm

```{r 7d}
a$ELEVATION <- round(a$ELEVATION, 1)

```


7e. show the first five lines of the data.frame

```{r 7e}
head(a, n=5)
#or
a[1:5, ]
```



7f. Save "a" in a file on your disk (comma separated by values format)

```{r 7f}
write.csv(a, paste0(tempdir(),'\\test.csv'), row.names=FALSE)

```


8. Also read file "Tanzania-All-Waterpoints-July-2015.csv"

```{r 8}
w <- read.csv("./data/Tanzania-All-Waterpoints-July-2015.csv", stringsAsFactors=FALSE)

```


9. For each region, compute the total number of water points, and the total number of people served (POPULATION.SERVED),  and the most common quantity ("WATER_QUAN") and type ("WATERPOINT") (either use the "modal" function from the raster package; or  use the table function

```{r 9}
wa <- aggregate(w[, 'POPULATION.SERVED'], w[, 'REGION', drop=FALSE], length) 
colnames(wa)[2] <- 'nwp'
wb <- aggregate(w[, 'POPULATION.SERVED', drop=FALSE], w[, 'REGION', drop=FALSE], sum, na.rm=TRUE) 

library(raster)
wc <- aggregate(w[, c('WATER_QUAN', 'WATERPOINT')], w[, 'REGION', drop=FALSE], modal, na.rm=TRUE) 
# or
f <- function(x) names(which.max(table(x)))
wc <- aggregate(w[, c('WATER_QUAN', 'WATERPOINT')], w[, 'REGION', drop=FALSE], f) 

#combine:
water <- merge(wa, wb, by='REGION')
water <- merge(water, wc, by='REGION')
head(water)
```


10. Make a table that shows counts of water quality types ("WATER_QUAL") by region

```{r 10}
qual <- table(w$REGION, w$WATER_QUAL)
head(qual, n=2)
```


11. Make a new data.frame that combines the REGION level totals/averages for the  Health facilities and for the water points, and for the water quality to get something like:

| REGION | ZONE               | elevation | n   | nwp  | POPULATION.SERVED | WATER_QUAN   | WATERPOINT         | coloured | fluoride | fluoride abandoned | good | milky | salty | salty abandoned | soft | unknown |
|--------|--------------------|-----------|-----|------|-------------------|--------------|--------------------|----------|----------|--------------------|------|-------|-------|-----------------|------|---------|
| Arusha | Northern           | 1542.9    | 334 | 4564 | 1776454           | insufficient | communal standpipe | 18       | 161      | 1                  | 0    | 5     | 20    | 1               | 4227 | 131     |
| Dodoma | Central            | 1059.4    | 369 | 3532 | 4239573           | enough       | communal standpipe | 38       | 0        | 0                  | 0    | 21    | 594   | 22              | 2694 | 163     |
| Geita  | Lake               | 1721.7    | 143 | 1944 | 384               | enough       | hand pump          | 32       | 1        | 0                  | 0    | 33    | 51    | 1               | 1561 | 265     |
| Iringa | Southern Highlands | 1541.1    | 241 | 2292 | 108092            | enough       | communal standpipe | 3        | 0        | 0                  | 0    | 0     | 27    | 4               | 2191 | 67      |
| Kagera | Lake               | 1426.5    | 298 | 4531 | 572969            | enough       | communal standpipe | 64       | 5        | 0                  | 0    | 76    | 117   | 2               | 4066 | 201     |
| Katavi | Southern Highlands | 1162.5    | 77  | 1069 | 217               | enough       | hand pump          | 6        | 0        | 0                  | 0    | 6     | 127   | 2               | 907  | 21      |


```{r 11}
m <- merge(a, water, by='REGION')
q <- as.data.frame.matrix(qual)
m <- merge(m, q, by.x='REGION', by.y=0)
# or
m <- merge(a, water, by='REGION')
qq <- as.data.frame(qual)
qq <- reshape(qq, direction='wide', timevar='Var2', idvar='Var1')
m <- merge(m, qq, by.x='REGION', by.y=1)

```

