---
title: "Manipulating data in R"
author: "Benson Kenduiywo"
date: "26/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Reading data

Our first exercise entails reading data from files and summarizing and combining the records. This is very common and important step; but somewhat tricky to learn.  We will use the data in a file named "Operating-Health-Facilities-with-location-2014.csv" under "data" read the data in the file into a data.frame and answer all the questions by using R code only. 

Replace the ___________ with the appropriate code 

Feel free to use materials of Introduction to R in: http://rspatial.org/intr/.

1. Read the file into a data.frame.

```{r d1}
d <- read.csv("Operating-Health-Facilities-with-location-2014.csv", stringsAsFactors=FALSE)

```

2. how many rows and columns are there?

```{r d2}
dim(d)
nrow(d)
ncol(d)
```


3. what are the variable names?

```{r d3}
colnames(d)
head(d)
tail(d)
```


4. What is the number of facilities for each zone?
See [data exploration](https://rspatial.org/intr/7-explore.html#data-exploration)

```{r d4}
table(d$ZONE) 
summary(d)
plot(sort(d$LATITUDE))

```

5. Which region has most facilities (and how many), and which has least (and how many)?

```{r d4}
x <- table(d$REGION)
names(which.min(x))
min(x)
names(which.max(x))
max(x)

```



