---
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls(all=TRUE))	#Clears R memory
unlink(".RData") 

```

## Exploratory data analysis

It is important prior to starting any analysis to inspect data This can be done to check for any departures from normality or any existing trend. Principle plots like histogram, boxplots and measures of central tendency and dispersion can help detect distributions of data that deviates from the normal. In additions maps can aid illustrate any spatial trends.

### Load data

We will mostly use the meuse dataset available in **sp** package. The meuse data set provided by package sp is a data set comprising of four heavy metals measured in the top soil in a flood plain along the river Meuse, along with a handful of covariates. The process governing heavy metal distribution seems that polluted sediment is carried by the river, and mostly deposited close to the river bank, and areas with low elevation. This document shows a Geostatistical analysis of this data set. The data set was introduced by Burrough and McDonnell, 1998.

```{r d1, message=FALSE}
rm(list=ls(all=TRUE)) # Clear memory
#Load existing/install missing libraries
if (!require("pacman")) install.packages("pacman")
pacman::p_load(gstat,sp)
data(meuse)
df <- meuse
head(df, n=3)

```

### Measures of Cental Tendency

Measures of of center of tendency include: mean. median and mode. For example, using the topsoil copper concentration data in Meuses dataset these measures can be computed as follows in R.

#### Mean and Median

See below.

```{r m1}
vMean   <- mean(df$copper)
cat("The mean is ", vMean)
vMedian <- median(df$copper)
cat("The median is ", vMedian)
```

#### Mode

There are several ways of finding the mode. We will illustrate here using a designed function and an inbuilt function in modeest library.

```{r m2, message=FALSE}
pacman::p_load(modeest)
attach(df)
getmode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}
vMode <- getmode(copper)
cat("The mode using the designed function is ", vMode)
vMode <- mlv(copper, method = "mfv")
str(vMode) #Chec structure
typeof(vMode) #Check variable type
cat("The mode using modeest package is ", vMode)

```


### Measures of Dispersion

Measures of dispersion include: variance, standard deviation, kurtosis, coefficient of variation and kurtosis. To illustrate this we still use meuse data in sp package.

#### Variance and standard deviation

```{r m3, message=FALSE}
data(meuse)
attach(meuse)
#compute variance of copper
var(copper)
#Standard deviation of copper 
sd(copper)
#Or You can compare the two approaches as below:
sd(copper)==sqrt(var(copper))

```

#### Kurtosis and skewness

Intuitively, the kurtosis describes the tail shape of the data distribution. The normal distribution has zero kurtosis and thus the standard tail shape. For instance the copper data is more peaked than the normal distribution since the kurtosis > 0. This is also illustrated by a histogram plot overlaid with the normal distribution curve.

On the other hand, skewness is a measure of symmetry. As a rule, negative skewness indicates that the mean of the data values is less than the median, and the data distribution is left-skewed. Positive skewness would indicate that the mean of the data values is larger than the median, and the data distribution is right-skewed.

```{r m4, message=FALSE}
pacman::p_load(e1071)
kurtosis(copper)
hist(copper, prob=TRUE,breaks=20,main="Normal distribution curve over histogram", xlab= "Copper")
curve(dnorm(x, mean=mean(copper), sd=sd(copper)), add=TRUE)
abline(v=mean(copper), col="red")
text(mean(copper),0.04,"Mean", col = "red", adj = c(0, -.1))
abline(v=median(copper), col="blue")
text(median(copper),0.04,"Median", col = "blue", adj = c(1, -.1))
cat("Skewness of copper is ", skewness(copper), " . This means that the data is positively skewed as illustrated on the histogram plot.")
```


### Normal Distribution

Here we illustrate how normal distribution looks like. In R you can generate a the normal distribution using a specified mean and standard deviation. 

```{r m5}
x<-rnorm(100000,mean=10, sd=2) 
hist(x,breaks=150,xlim=c(0,20),freq=FALSE)
abline(v=10, lwd=5)
abline(v=c(4,6,8,12,14,16), lwd=3,lty=3)
```

### Exploring spatial trend

Spatial exploratory data analysis starts with the plotting of maps with a
measured variable.

```{r sp1}
coordinates(meuse) <- c("x", "y")
spplot(meuse, "zinc", do.log = T)
bubble(meuse, "zinc", do.log = T, key.space = "bottom")

```

The evident structure here is that zinc concentration is larger close to the
river Meuse banks.

In case of an evident spatial trend, such as the relation between top soil zinc concentration and distance ($\text{zinc} \propto \beta_0 + \beta_1(\text{dist})$) to the river, we make a plot to illustrate this relationship.


```{r sp2, message=FALSE}
pacman::p_load(lattice)
xyplot(log(zinc) ~ sqrt(dist), as.data.frame(meuse))

```

The plot shows that Zinc concentration linearly decreases with increasing distance from the river. This supports the former premise that heavy metal concentration is higher along the river banks.

We can also model this relationship using a linear regression and plot maps with fitted values and with residuals.

```{r sp3}
zn.lm <- lm(log(zinc) ~ sqrt(dist), meuse)
meuse$fitted.s <- predict(zn.lm, meuse) - mean(predict(zn.lm, meuse))
meuse$residuals <- residuals(zn.lm)
spplot(meuse, c("fitted.s", "residuals"))

```

These plots reveal that although the trend removes a large part of the variability, the residuals do not appear to behave as spatially unstructured (spatially independent) or white noise. We see that **residuals** with a similar value occur regularly close to another. This indicates some degree of **spatial dependency** with distance. More analysis will
take place when we further analyse these data in the context of geostatistical
models; first we will go through simple, **non-geostatistical interpolation** approaches.

Let us first look at another example before we go to **non-geostatistical interpolation**. Here we use data from [Piikki et al.] 2019(https://www.mdpi.com/2071-1050/11/15/4038).

```{r t1}
d <- read.csv("./data/soil_data_CIAT.csv", stringsAsFactors=FALSE)
head(d, n=3)

```

Examine the spatial trend of the variables (Bulk density, Carbon, Sand and Clay).

```{r t2, out.width="100%"}
pacman::p_load(gridExtra)
coordinates(d) <- c("X", "Y")
plots = lapply(names(d)[2:5], function(.x) spplot(d,.x,do.log = T, main=.x))
do.call(grid.arrange, plots)

```

Its clear that there is some spatial structure in these variables. This is also supported by the trend depicted their residuals obtained by regressing the variables over their locations/coordinates:

```{r t2a}
#Carbon
ca_lm <- lm(Carbon ~ X + Y, d)
d$fitted.ca <- predict(ca_lm, d) - mean(predict(ca_lm, d))
d$residuals.ca <- residuals(ca_lm)
#Bulk density
bd_lm <- lm(Bulk_density ~ X + Y, d)
d$fitted.bd <- predict(bd_lm, d) - mean(predict(bd_lm, d))
d$residuals.bd <- residuals(bd_lm)
#sand
sa_lm <- lm(Sand ~ X + Y, d)
d$fitted.sa <- predict(sa_lm, d) - mean(predict(sa_lm, d))
d$residuals.sa <- residuals(sa_lm)
spplot(d, names(d)[6:11])
#plots = lapply(names(d)[6:11], function(.x) spplot(d,.x, main=.x))
#do.call(grid.arrange, plots)

```

Is there any relationship between organic carbon with clay and sand?

```{r t3}
par(mfrow=c(1,2))
plot(Carbon ~ Clay + Sand , d)

```

We have seen how to explore data using plots and some measures. Next lesson illustrates how to analyze the data using **non-geostatistical interpolation**. 

### References

Piikki, K.; Söderström, M.; Sommer, R.; Da Silva, M.; Munialo, S.; Abera, W. [A Boundary Plane Approach to Map Hotspots for Achievable Soil Carbon Sequestration and Soil Fertility Improvement.](https://www.mdpi.com/2071-1050/11/15/4038) *Sustainability* 2019, 11, 4038.
